---
title: "RNN Genitive Alternation"
output: html_notebook
---

Do RNNs show evidence for humanlike preferences in the genitive alternation? Specifically, do they show a preference for the NP that is more animate, definite, and short to come earlier? Are they sensitive to biases due to the semantic classes of the genitive relationship?

```{r}
rm(list = ls())
library(tidyverse)
library(lme4)
library(lmerTest)
library(plotrix)
library(stringr)
library(readxl)

S_REGION_ORDER = c("Intro", "Def Possessor", "Indef Possessor", "S", "Possessum", "End")
OF_REGION_ORDER = c("Intro", "the", "Possessum", "of", "Def Possessor", "Indef Possessor", "End")

rename_models = function(d) {
  d %>% 
    mutate(model=if_else(model == "gulordava", "GRNN",
                 if_else(model == "google", "JRNN",
                 if_else(model == "kenlm", "n-gram", model))))
    
}

remove_na = function(x) {
  x[!is.na(x)]
}

remove_prefix = function(s) {
  sapply(s, function(s) {
    str_split(s, "-")[[1]][2]
  })
}

items = read_excel("genitive-items.xlsx") %>%
  mutate(sent_index=seq(0, 15)) %>%
  rename(possessor_animacy=`Possessor Animate?`,
         possessum_animacy=`Possessum Animate?`) %>%
  mutate(possessor_animacy=if_else(possessor_animacy, "anim", "inanim"),
         possessum_animacy=if_else(possessum_animacy, "anim", "inanim"))

d = read_csv("tests/combined_results.csv") %>%
  select(-1, -2) %>%
  mutate(unk=unk == "True") %>%
  separate(condition, sep="_", into=c("construction", "possessor_definiteness", "possessor_length")) %>%
  inner_join(items)

d_agg = d %>% 
  group_by(model, sent_index, construction, possessor_definiteness, possessor_length, possessor_animacy, possessum_animacy) %>% 
    summarise(surprisal=sum(surprisal),
              unk=any(unk)) %>%
    ungroup() %>% 
  filter(!unk) %>%
  mutate(construction_numeric=if_else(construction == "of", -1, 1),
         possessor_length=factor(possessor_length, levels=c("short", "long")),
         possessor_length_numeric=if_else(possessor_length == "short", -1, 1),
         possessor_definiteness=factor(possessor_definiteness, levels=c("indef", "def")),
         possessor_definiteness_numeric=if_else(possessor_definiteness=="indef", -1, 1),
         possessor_animacy=factor(possessor_animacy, levels=c("inanim", "anim")),
         possessor_animacy_numeric=if_else(possessor_animacy == "inanim", -1, 1),
         possessum_animacy=factor(possessum_animacy, levels=c("inanim", "anim")),
         possessum_animacy_numeric=if_else(possessum_animacy == "inanim", -1, 1))
                                  

h = read_csv("human_ratings.csv") %>%
  rename(sent_index=item) %>%
  inner_join(select(items, sent_index, possessor_animacy, possessum_animacy)) %>%
  mutate(construction_numeric=if_else(construction == "of", -1, 1),
         possessor_length=factor(possessor_length, levels=c("short", "long")),
         possessor_length_numeric=if_else(possessor_length == "short", -1, 1),
         possessor_definiteness=factor(possessor_definiteness, levels=c("indef", "def")),
         possessor_definiteness_numeric=if_else(possessor_definiteness=="indef", -1, 1),
         possessor_animacy=factor(possessor_animacy, levels=c("inanim", "anim")),
         possessor_animacy_numeric=if_else(possessor_animacy == "inanim", -1, 1),
         possessum_animacy=factor(possessum_animacy, levels=c("inanim", "anim")),
         possessum_animacy_numeric=if_else(possessum_animacy == "inanim", -1, 1))
  
```

In this regression, a positive interaction with construction indicates that the of-construction is favored, and a negative interaction with construction indicates that the s-construction is favored. (Logic: s is coded 1, so a negative effect means that surprisal for s decreases. of is coded -1, so a positive effect means that surprisal for of decreases.)

```{r}

m_google = lmer(surprisal ~ construction_numeric * 
                  (possessor_length_numeric + possessor_definiteness_numeric + 
                     possessor_animacy_numeric + possessum_animacy_numeric) +
                    (construction_numeric * (possessor_length_numeric + possessor_definiteness_numeric)|sent_index),
           data=filter(d_agg, model == "google"))
summary(m_google)

m_gul =  lmer(surprisal ~ construction_numeric * 
                  (possessor_length_numeric + possessor_definiteness_numeric + 
                     possessor_animacy_numeric + possessum_animacy_numeric) +
                    (construction_numeric * (possessor_length_numeric + possessor_definiteness_numeric)|sent_index),
           data=filter(d_agg, model == "gulordava"))
summary(m_gul)

m_kenlm = lmer(surprisal ~ construction_numeric * 
                  (possessor_length_numeric + possessor_definiteness_numeric + 
                     possessor_animacy_numeric + possessum_animacy_numeric) +
                    (construction_numeric * (possessor_length_numeric + possessor_definiteness_numeric)|sent_index),
           data=filter(d_agg, model == "kenlm"))
summary(m_kenlm)

m_google_kenlm = lmer(surprisal ~ model * construction_numeric * 
                  (possessor_length_numeric + possessor_definiteness_numeric + 
                     possessor_animacy_numeric + possessum_animacy_numeric) +
                    (model * construction_numeric * (possessor_length_numeric + possessor_definiteness_numeric)|sent_index),
           data=filter(d_agg, model == "google" | model == "kenlm"))
summary(m_google_kenlm)

m_gulordava_kenlm = lmer(surprisal ~ model * construction_numeric * 
                  (possessor_length_numeric + possessor_definiteness_numeric + 
                     possessor_animacy_numeric + possessum_animacy_numeric) +
                    (model * construction_numeric * (possessor_length_numeric + possessor_definiteness_numeric)|sent_index),
           data=filter(d_agg, model == "gulordava" | model == "kenlm"))
summary(m_gulordava_kenlm)

m_human = lmer(acceptability ~ 
            construction_numeric * (possessor_length + possessor_definiteness + possessor_animacy + possessum_animacy) +
           (construction_numeric * (possessor_length + possessor_definiteness + possessor_animacy + possessum_animacy)|sent_index) +
           (construction_numeric * (possessor_length + possessor_definiteness + possessor_animacy + possessum_animacy)|WorkerId),
           data=h)
summary(m_human)

```

All as expected in the Gulordava LSTM, though not everything is significant.
All is as expected in the Google LSTM, except the effect of possessor definiteness is in the wrong direction! (but ns)

Let's visualize.

```{r}

d_effect = d_agg %>%
  select(-construction_numeric) %>% 
  spread(construction, surprisal) %>% 
  mutate(of_preference=s - of)

d_by_animacy = d_effect %>%
  group_by(model, sent_index) %>%
    mutate(relevant_of_preference=of_preference - mean(of_preference)) %>%
    ungroup() %>%
  group_by(model, possessor_animacy, possessum_animacy) %>%
    summarise(m_of_preference=mean(of_preference),
              s=std.error(relevant_of_preference),
              lower=m_of_preference-1.96*s,
              upper=m_of_preference+1.96*s) %>%
    ungroup() 

h_by_animacy = h %>%
  group_by(sent_index) %>%
    mutate(relevant_acceptability=acceptability - mean(acceptability, na.rm=T)) %>%
    ungroup() %>%
  group_by(WorkerId) %>%
    mutate(relevant_acceptability=relevant_acceptability - mean(relevant_acceptability, na.rm=T)) %>%
    ungroup() %>%
  group_by(construction, possessor_animacy, possessum_animacy) %>%
    summarise(n=n(),
              m=mean(acceptability,na.rm=T),
              sd=sd(relevant_acceptability, na.rm=T)) %>%
    ungroup() %>%
  unite(msn, m, sd, n) %>%
  spread(construction, msn) %>%
  separate(s, into=c("m_s", "sd_s", "n_s"), sep="_") %>%
  separate(of, into=c("m_of", "sd_of", "n_of"), sep="_") %>%
  mutate(m_s=as.numeric(m_s),
         m_of=as.numeric(m_of),
         sd_s=as.numeric(sd_s),
         sd_of=as.numeric(sd_of),
         n_s=as.numeric(n_s),
         n_of=as.numeric(n_of)) %>%
  mutate(m_of_preference = m_of - m_s,
         s=sqrt(sd_of^2/n_of + sd_s^2/n_s),
         upper=m_of_preference+1.96*s,
         lower=m_of_preference-1.96*s) %>%
  mutate(model="Human")
  
d_by_animacy %>%
  bind_rows(h_by_animacy) %>%
  rename_models() %>%
  ggplot(aes(x=possessum_animacy, fill=possessor_animacy, y=m_of_preference, ymin=lower, ymax=upper)) +
    geom_bar(stat="identity", position="dodge") +
    geom_errorbar(color="black", width=.5, position=position_dodge(width=.9)) +
    facet_wrap(~model,scale="free_y") +
    labs(x="Possessum animacy", y="Preference for of-construction", fill="Possessor animacy") +
    theme(legend.position="bottom")

ggsave("genitive_alternation_animacy.pdf", width=6, height=5)
```

Basically we favor the s-genitive with an animate posessor (as expected), and favor the of-genitive very slightly with an animate possessum in Google and Kenlm but not Gulordava.

Let's look for animacy effects in regressions.

```{r}
m_google = lmer(surprisal ~ construction_numeric * (possessor_animacy + possessum_animacy) +
                    (construction_numeric * (possessor_animacy + possessum_animacy)|sent_index),
           data=filter(d_agg, model == "google"))
summary(m_google)

m_gul = lmer(surprisal ~ construction_numeric * (possessor_animacy + possessum_animacy)
                 +  (1|sent_index),
           data=filter(d_agg, model == "gulordava"))
summary(m_gul)
```

Nothing anywhere... possibly the between-item design is too weak.

Now what about by definiteness and length?

```{r}
d_by_def_length = d_effect %>%
  group_by(model, sent_index) %>%
    mutate(relevant_of_preference = of_preference - mean(of_preference)) %>%
    ungroup() %>%
  group_by(model, possessor_definiteness, possessor_length) %>%
    summarise(m_of_preference=mean(of_preference),
              s=std.error(relevant_of_preference),
              lower=m_of_preference-1.96*s,
              upper=m_of_preference+1.96*s) %>%
    ungroup() 

h_by_def_length = h %>%
  group_by(sent_index) %>%
    mutate(relevant_acceptability=acceptability - mean(acceptability, na.rm=T)) %>%
    ungroup() %>%
  group_by(WorkerId) %>%
    mutate(relevant_acceptability=relevant_acceptability - mean(relevant_acceptability, na.rm=T)) %>%
    ungroup() %>%
  group_by(construction, possessor_definiteness, possessor_length) %>%
    summarise(n=n(),
              m=mean(acceptability,na.rm=T),
              sd=sd(relevant_acceptability,na.rm=T)) %>%
    ungroup() %>%
  unite(msn, m, sd, n) %>%
  spread(construction, msn) %>%
  separate(s, into=c("m_s", "sd_s", "n_s"), sep="_") %>%
  separate(of, into=c("m_of", "sd_of", "n_of"), sep="_") %>%
  mutate(m_s=as.numeric(m_s),
         m_of=as.numeric(m_of),
         sd_s=as.numeric(sd_s),
         sd_of=as.numeric(sd_of),
         n_s=as.numeric(n_s),
         n_of=as.numeric(n_of)) %>%
  mutate(m_of_preference = m_of - m_s,
         s=sqrt(sd_of^2/n_of + sd_s^2/n_s),
         upper=m_of_preference+1.96*s,
         lower=m_of_preference-1.96*s) %>%
  mutate(model="Human")


d_by_def_length %>%
  bind_rows(h_by_def_length) %>%
  rename_models() %>%
  ggplot(aes(x=possessor_length, fill=possessor_definiteness, y=m_of_preference, ymin=lower, ymax=upper)) +
    geom_bar(stat="identity", position="dodge") +
    geom_errorbar(color="black", width=.5, position=position_dodge(width=.9)) +
    facet_wrap(~model, scale="free_y") +
    labs(x="Possessor length", fill="Possessor definiteness", y="Preference for of-construction") +
    theme(legend.position="bottom")

ggsave("genitive_alternation_etc.pdf", width=6, height=5)
```

Long possessor strongly pulls for the of-genitive. There seems to be a preference for the s-genitive with a definite possessor when the possessor is short, but when the possessor is long this effect disappears or is reversed.

```{r}
m_google = lmer(surprisal ~ construction_numeric * (possessor_definiteness + possessor_length) +
                    (construction_numeric * (possessor_definiteness + possessor_length)|sent_index),
           data=filter(d_agg, model == "google"))
summary(m_google)

m_gul = lmer(surprisal ~ construction_numeric * (possessor_definiteness + possessor_length)
                 +  (construction_numeric*(possessor_definiteness + possessor_length)|sent_index),
           data=filter(d_agg, model == "gulordava"))
summary(m_gul)

m_kenlm = lmer(surprisal ~ construction_numeric * (possessor_definiteness + possessor_length)
                 +  (construction_numeric*(possessor_definiteness + possessor_length)|sent_index),
           data=filter(d_agg, model == "kenlm"))
summary(m_kenlm)
```

Length effect is significant; definiteness effect is not.

A particularly interesting case is when the possessor is long and definite or indefinite, because then the model can't rely on n-gram statistics and must carry forward a representation of the definiteness of the possessor. This analysis was not pre-registered, but let's restrict our attention to the long possessor case, and see if there is an effect of definiteness there.

```{r}
m_google = lmer(surprisal ~ construction_numeric * possessor_definiteness +
                    (1|sent_index),
           data=filter(d_agg, model == "google", possessor_length == "long"))
summary(m_google)

m_gul = lmer(surprisal ~ construction_numeric * (possessor_definiteness)
                 +  (1|sent_index),
           data=filter(d_agg, model == "gulordava", possessor_length == "long"))
summary(m_gul)
```

Alas, none of them get it!