---
title: "RNN Particle Shift II"
output: html_notebook
---

Do RNNs show evidence for humanlike preferences in the genitive alternation? Specifically, do they show a preference for the NP that is more animate, definite, and short to come earlier? Are they sensitive to biases due to the semantic classes of the genitive relationship?

For consistency with the Heavy NP Shift analysis, "shifted" refers to the order V P NP; "unshifted" refers to V NP P. 

```{r}
rm(list = ls())
library(tidyverse)
library(lme4)
library(lmerTest)
library(plotrix)
library(stringr)
library(readxl)

remove_na = function(x) {
  x[!is.na(x)]
}

remove_prefix = function(s) {
  sapply(s, function(s) {
    str_split(s, "-")[[1]][2]
  })
}

rename_models = function(d) {
  d %>% 
    mutate(model=if_else(model == "gulordava", "GRNN",
                 if_else(model == "google", "JRNN",
                 if_else(model == "kenlm", "n-gram", "UNKNOWN MODEL!"))))
    
}

items = read_excel("particle-shift.xlsx") %>%
  mutate(sent_index=seq(1, 32)) %>%
  rename(animate=`Animate?`) %>%
  mutate(animate=if_else(animate == "TRUE", "anim", "inanim"))
         

d = read_csv("tests/combined_results.csv") %>%
  select(-1, -2) %>%
  mutate(unk=unk == "True") %>%
  separate(condition, sep="_", into=c("shifted", "length")) %>%
  inner_join(items)

  
d_agg = d %>% 
  group_by(model, sent_index, shifted, length, animate) %>% 
    summarise(surprisal=sum(surprisal),
              unk=any(unk)) %>%
    ungroup() %>% 
  mutate(shifted=if_else(shifted == "shifted", "unshifted", "shifted")) %>% # Standardized shifting convention
  mutate(shifted_numeric=if_else(shifted == "shifted", 1, -1),
         length=factor(length, levels=c("short", "long")),
         animate=factor(animate, levels=c("inanim", "anim")),
         shifted=factor(shifted, levels=c("unshifted", "shifted")))

bads = d_agg %>%
  group_by(model, sent_index) %>%
    summarise(unk=any(unk)) %>%
    ungroup() %>%
  filter(!unk)

d_agg = inner_join(d_agg, bads)
                       
h = read_csv("human_data.csv") %>%
    rename(animate=animacy,
           shifted=order) %>%
    mutate(animate=factor(animate, levels=c("inanim", "anim")),
           length=factor(length, levels=c("short", "long"))) %>%
    separate(item, into=c("item", "junk", "trash"), sep="_") %>%
    select(-junk, -trash)
```

# Regressions 

```{r}

m_google = lmer(surprisal ~ shifted * (length + animate) + (shifted+length|sent_index),
           data=filter(d_agg, model == "google"))
summary(m_google) # Works as expected for length; animacy ns

m_gul = lmer(surprisal ~ shifted * (length+animate) + (shifted+length|sent_index),
           data=filter(d_agg, model == "gulordava"))
summary(m_gul) # Same

m_kenlm = lmer(surprisal ~ shifted * (length+animate) + (shifted+length|sent_index),
           data=filter(d_agg, model == "kenlm"))
summary(m_kenlm) # ns all around!

m_google_kenlm = lmer(surprisal ~ shifted * (length+animate) * model + (model*(shifted+length)|sent_index),
           data=filter(d_agg, model == "google" | model == "kenlm"))
summary(m_google_kenlm)

m_gulordava_kenlm = lmer(surprisal ~ shifted * (length+animate) * model + (model*(shifted+length)|sent_index),
           data=filter(d_agg, model == "gulordava" | model == "kenlm"))
summary(m_gulordava_kenlm)

m_google_gulordava = lmer(surprisal ~ shifted * (length+animate) * model + (model*(shifted+length)|sent_index),
           data=filter(d_agg, model == "gulordava" | model == "google"))
summary(m_google_gulordava)

m_human = lmer(acceptability ~ shifted * animate * length + 
                 (shifted * (animate * length)|item) +
                 (shifted * (animate * length)|WorkerId),
           data=h)
summary(m_human) 

```

Let's visualize. Use the "length penalty" concept from the heavy NP shift analysis.

```{r}

d_agg %>%
  spread(length, surprisal) %>%
  mutate(length_penalty = long - short) %>%
  group_by(model, shifted, animate) %>%
    summarise(m=mean(length_penalty),
              s=std.error(length_penalty),
              upper=m+1.96*s,
              lower=m-1.96*s) %>%
    ungroup() %>%
  rename_models() %>%
  mutate(shifted=if_else(shifted == "shifted", "V-P-NP", "V-NP-P")) %>%
  ggplot(aes(x=animate, y=m, fill=shifted, ymin=lower, ymax=upper)) +
    geom_bar(stat="identity", position="dodge") +
    geom_errorbar(width=.5, position=position_dodge(width=.9), color="black") +
    facet_wrap(~model) +
    labs(fill="Order", x="NP animacy", y="Surprisal penalty for long NP (bits)")

```

We'll compare to human acceptability ratings; first let's process this stuff into an overall shifted-preference.

```{r}

d_agg_pref = d_agg %>%
  select(-shifted_numeric) %>%
  spread(shifted, surprisal) %>%
  mutate(shifted_preference = unshifted - shifted) %>%
  group_by(model, sent_index) %>%
    mutate(relevant_shifted_preference = shifted_preference - mean(shifted_preference)) %>%
    ungroup() %>%
  group_by(model, length, animate) %>%
    summarise(m=mean(shifted_preference),
              s=std.error(relevant_shifted_preference),
              upper=m+1.96*s,
              lower=m-1.96*s) %>%
    ungroup() %>%
  rename_models() 

d_agg_pref %>%
  ggplot(aes(x=length, y=m, fill=animate, ymin=lower, ymax=upper)) +
    geom_bar(stat="identity", position="dodge") +
    geom_errorbar(width=.5, position=position_dodge(width=.9), color="black") +
    facet_wrap(~model) +
    labs(fill="NP Animacy", x="NP Length", y="Preference for order V-P-NP (bits)")

ggsave("particle_shift.pdf", width=7, height=5)

```

Now add in the human data, formatting it so it can be joined in with the summarised model data.

```{r}

h_by_length_animacy = h %>%
  group_by(item) %>%
    mutate(relevant_acceptability=acceptability - mean(acceptability, na.rm=T)) %>%
    ungroup() %>%
  group_by(WorkerId) %>%
    mutate(relevant_acceptability=relevant_acceptability - mean(relevant_acceptability, na.rm=T)) %>%
    ungroup() %>%
    
  group_by(shifted, animate, length) %>%
    summarise(n=n(),
              m=mean(acceptability, na.rm=T),
              sd=sd(acceptability, na.rm=T)) %>%
    ungroup() %>%
  unite(msn, m, sd, n) %>%
  spread(shifted, msn) %>%
  separate(shifted, into=c("m_shifted", "sd_shifted", "n_shifted"), sep="_") %>%
  separate(unshifted, into=c("m_unshifted", "sd_unshifted", "n_unshifted"), sep="_") %>%
  mutate(m_shifted=as.numeric(m_shifted),
         m_unshifted=as.numeric(m_unshifted),
         sd_shifted=as.numeric(sd_shifted),
         sd_unshifted=as.numeric(sd_unshifted),
         n_shifted=as.numeric(n_shifted),
         n_unshifted=as.numeric(n_unshifted)) %>%
  mutate(m=m_shifted - m_unshifted,
         s=sqrt(sd_shifted^2/n_shifted + sd_unshifted^2/n_unshifted),
         upper=m+1.96*s,
         lower=m-1.96*s) %>%
  mutate(model="Human")


d_agg_pref %>%
  bind_rows(h_by_length_animacy) %>%
  ggplot(aes(x=length, y=m, fill=animate, ymin=lower, ymax=upper)) +
    geom_bar(stat="identity", position="dodge") +
    geom_errorbar(width=.5, position=position_dodge(width=.9), color="black") +
    facet_wrap(~model, scales="free_y") +
    labs(fill="NP Animacy", x="NP Length", y="Preference for order V-Part-NP") + 
    theme(legend.position="bottom")
    
ggsave("particle_shift.pdf", width=6, height=5)
```